---
title: "Cost of Living in San Francisco "
about: 
  template: solana
---

### USING R TO UNDERSTAND COST OF LIVING IN SAN FRANCISCO

The goal of this project is to understand the rental market in San Francisco, California. This project uses the 2020 5-Year American Community Survey (ACS) and the 2024 5-Year ACS. This project uses tidycensus, tidyverse, tmap, and other R libraries to analyze the ACS data.

### GETTING THE LIBRARIES & DATA

The first step of this analysis is to install the necessary libraries to acquire the San Francisco data.

```{r}
#| message: false
#| code-fold: true

library(tidycensus) 
library(sf)
library(tidyverse)
library(spdep)
library(tmap)
library(mapview)

```

```{r}
#| message: false
#| code-fold: true

# setting the variables 
sf_var <- c(
  med_gross_rent = "B25064_001",
  med_income = "B19013_001",
  med_housing_cost = "B25105_001")

```

```{r}
#| message: false
#| code-fold: true
#| results: hide

# SF data
sf_data <- get_acs(
  geography = "tract", 
  variables = sf_var,
  state = "CA", 
  county = "San Francisco", 
  output = "wide",
  geometry = TRUE, 
  year = 2020
) %>%
  st_transform(7131)|> # filtering out islands
  filter(!GEOID %in% c(
    "06075980401",  # Farallon Islands
    "06075017902",  # Treasure Island / Yerba Buena
    "06075017901",  # Treasure Island
    "06075017903",  # Treasure Island 
    "06075980300"   # water tract
  )) |>
  filter(!st_is_empty(geometry)) |>
  filter(st_is_valid(geometry))



```

### UNDERSTANDING RENT 

This figure shows the median gross rent in San Francisco. Overall, rent in San Francisco is high, but this figure indicates the range across the central city.

```{r}
#| message: false
#| code-fold: true

# mapping median rent estimate 
med_rent_map <- ggplot(sf_data, aes(fill = med_gross_rentE)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = "PuBuGn", 
                     direction = 1,
                     labels = scales::label_dollar()) +
  theme_void() +
  labs(title = "Median Gross Rent in San Francisco, CA",
       fill = "Median Gross Rent",
       caption = "Source: 2020 5-Year ACS, US Census Bureau") +
  theme(text = element_text(family = ""), 
        plot.title = element_text(hjust = 0.5)) 


med_rent_map

```

### STATISTICAL ANALYSIS 

Using the Local Gi\*, we are able to understand the clusters of high and low rent in the region.

```{r}
#| message: false
#| code-fold: true
#| results: hide

# mapping clusters of high low and high rent in San Francisco 
# weights matrix

sf_data <- sf_data |> filter(!is.na(med_gross_rentE))

neighbors <- poly2nb(sf_data, queen = TRUE)
summary(neighbors)

sf_data[card(neighbors) == 0, ] |> select(GEOID, NAME)

weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
weights$weights[[1]]

neighbors[[1]]

sf_data$lag_estimate <- lag.listw(weights, sf_data$med_gross_rentE)


# For Gi*, re-compute the weights with `include.self()`
localg_weights <- nb2listw(include.self(neighbors))

sf_data$localG <- as.numeric(localG(sf_data$med_gross_rentE, localg_weights))

```

This map highlights statistically significant clusters of high and low rent. The red, or high values, indicate tracts with high median gross rent surrounded by other tracts with high median gross rent. Low values are shown in blue, which indicate tracts with low median gross rent surrounded by other low median gross rent tracts.

```{r}
#| message: false
#| code-fold: true
#| fig-height: 6



# Map of Local Gi scores w/spatial clusters highlighted
sf_data <- sf_data %>%
mutate(hotspot = case_when(
localG >= 2.56 ~ "High cluster",
localG <= -2.56 ~ "Low cluster",
TRUE ~ "Not significant"
))

tmap_mode("view")

tm_shape(sf_data) +
  tm_polygons(
    fill = "hotspot",
    fill.scale = tm_scale_categorical(
      values = c("High cluster" = "red", "Low cluster" = "blue", "Not significant" = NA)
    ),
    fill.legend = tm_legend(title = "Rental Clusters"),
    col = NA,
    col_alpha = 0.5
  ) +
  tm_basemap("CartoDB.Positron") +
  tm_title("Areas with High & Low Rent Clusters in San Francisco, CA")

```

### UNDERSTANDING VACANCY 

Now that we have run visualizations and statistical tests to understand rent in San Francisco, I wanted to compare this with the vacancy rate of each census tract. This is done by acquiring the vacant housing and total housing units from the 2024 5-Year ACS.

```{r}
#| message: false
#| code-fold: true
#| results: hide

vacancy_vars <- c(vacant = "B25004_001", 
                  total = "B25002_001")

recent_vacancy <- get_acs(
  geography = "tract", 
  variables = vacancy_vars,
  state = "CA", 
  county = "San Francisco", 
  geometry = TRUE, 
  output = "wide", 
  year = 2024
) %>%
  st_transform(7131)|> # filtering out islands
  filter(!GEOID %in% c(
    "06075980401",  # Farallon Islands
    "06075017902",  # Treasure Island / Yerba Buena
    "06075017901",  # Treasure Island
    "06075017903",  # Treasure Island 
    "06075980300"   # water tract
  )) |>
  filter(!st_is_empty(geometry)) |>
  filter(st_is_valid(geometry)) 

sf_vacancy_rate <- recent_vacancy %>%
 mutate(vacancy_rate = vacantE/totalE)
 
```

This figure shows the vacancy rate across San Francisco.

```{r}
#| message: false
#| code-fold: true


# mapping vacancy rate
vacancy_map <- ggplot(sf_vacancy_rate, aes(fill = vacancy_rate)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = "BuPu", 
                     direction = 1) +
  theme_void() +
  labs(title = "Vacancy Rate in San Francisco, CA",
        fill = "Vacancy Rate",
       caption = "Source: 2024 5-Year ACS, US Census Bureau") + 
  theme(text = element_text(family = ""), 
        plot.title = element_text(hjust = 0.5)) 


vacancy_map

```

### PUTTING IT ALL TOGETHER

The final figure of this project is an interactive map with the vacancy rate, rental clusters, and rental cost. While this is not the full picture of the rental market in San Francisco, it shows cost, availability, and statistically significant regions of high and low rent across the city. The interactive map allows users to have vacancy rate or median gross rent shown, as well as the clusters of high and low median gross rent across the city.

```{r}
#| message: false
#| code-fold: true
#| fig-height: 8


# mapping vacancy rate
tmap_mode("view")

combined_map <- tm_shape(sf_data) +
  tm_polygons(
    fill = "med_gross_rentE",
    fill.scale = tm_scale_continuous(
      values = "PuBuGn",
      label.format = list(prefix = "$", big.mark = ",")
    ),
    fill.legend = tm_legend(title = "Median Gross Rent",
                            reverse = TRUE),
                             col = "grey") +
  tm_shape(sf_vacancy_rate |> filter(!is.na(vacancy_rate))) +
  tm_polygons(
    fill = "vacancy_rate",
    fill.scale = tm_scale_continuous(
      values = "brewer.BuPu"
    ),
    fill.legend = tm_legend(title = "Vacancy Rate", reverse = TRUE),
    col = "grey"
  ) +
  tm_shape(sf_data |> filter(hotspot != "Not significant")) +
  tm_borders(
    col = "hotspot",
    col.scale = tm_scale_categorical(
      values = c("High cluster" = "red", "Low cluster" = "blue")
    ),
    col.legend = tm_legend(title = "Rental Clusters"),
    lwd = 2
  ) + 
  tm_basemap("CartoDB.Positron") +
  tm_title("Vacant Housing Units in San Francisco, CA")

  
combined_map

```
